Control of Prosumer Networks

### 2016-02-17 Article.tex
update

Modernizing the power grids and increasing the share of renewables in the production are two substantial objectives of the energetic transition. Because of progresses in communication, data management, and storage, the upcoming emergence of a power grid "2.0", often called smart grid, appears as a cross-disciplinary challenge of the 21st century \cite{Ramchurn}.

It is believed that these new ways of producing and managing energy, will also impact how end users consume. More precisely, the consumer   of the future is expected to be more involved in the system operation. Using dynamic signals (real time electricity prices for instance), load curves could be shaped to some extend as to agree with the production conditions of the main grid.
Furthermore, the emergence of renewable distributed energy resources (DER) in the distribution networks, possibly hold by individuals,
will completely modify the today end user agent model.

These new agents that can both consume and produce are often called prosumers \cite{Rathnayaka2012}. This decentralized production is believed to lead to more resilient systems than today centralized power grid. Nonetheless, the difficulty of dealing with decentralized production is coupled with the high uncertainty of renewables. Ensure system stability in these conditions appears thus as a complex task.

Relying on the ability of the smart grid to obtain, communicate and analyse data, several solutions could  improve the stability of such complex systems. Among these solutions, demand side management and energy storage are two of the most popular ones. In this paper, we explore the use of storage in a network of prosumers whose power outputs are susceptible to change due to various external causes.

@@

We model this network as a coupled oscillator network, where each node is represented as an oscillator with a phase angle and a natural frequency that are impacted by the ones of their neighbors. Since the work of Kuramoto [], networks of coupled oscillators have been used to model various phenomenons such as fireflies, ex2, or power grids. Indeed, it is well-known that synchronization of all rotating elements to a common frequency is necessary within a power grid for stable operation. Inspired by \cite{Filatrella2008}, we use a second order Kuramoto model in order to obtain a simplified dynamics for the power grid (see Section \ref{sec:The_Model}).

We will see that under some conditions, the system synchronizes to a common frequency. However, this synchronized state is suceptible to break if perturbations occur. If, for example, the production of some generator decreases while the load remains the same, the imbalance between production and consumption might drive the system to unstable conditions. These kinds of perturbations are very likely in a scenario with high penetration of DER. Multiple control techniques, with different time scales, exits in order to cope with such events. Among others, it has been argued that storage and vehicle to grid (V2G) technology could be used for frequency regulation. If a sudden drop or raise of the frequency due to an imbalance is detected, some devices are charged or discharged as to restore synchrony. This approach poses a few questions such as the location of the storage devices, or the quantities of power that should be injected or absorbed. Since charging and discharging a battery is costly, we are particularly interested in the energy necessary to restaure synchrony. Moreover, in a prosumer scenario where generators and loads tend not to be fixed, we would like to find a subset of locations for installing storage such that the energy is low on average.

We will see that these storage actions can be interpreted as an additional term in the oscillators dynamics for a subset of nodes. We propose here to use complex network control theory \cite{Liu2015} in order to model the impact of the storage equipments on the network dynamics (see Section \ref{sec:control_and_submodularity}). Placing storage at some node can thus be seen as selecting this node in the driver set. Selecting the drivers that, on average, will control the network with low energy is a hard problem. Nevertheless, recent studies showed the existence of relationships between controllability and submodular set functions \cite{Summers2014}. These functions, that we introduce in more depth in section X, have the interesting property of having diminishing returns as we increase the size of the set considered. Although finding the optimal set is NP-hard, \cite{Minoux} proposed a simple greedy algorithm with a worst case guaranty.

@@

The condition for the power transmission between i and j is that the two devices do not operate in phase : The phase difference between i and j is : $ \delta_j(t) - \delta_i(t) = \Omega t + \theta_j(t) - \Omega t - \theta_i(t) = \theta_j(t) - \theta_i(t) $.

The transmitted power along the line can thus be written as :

@@

\begin{equation}
\begin{array}{lll}
P_{S, i}  =  I \Omega \ddot{ \theta_i }  + \left[ I \ddot{ \theta_i } + 2 K_D \Omega \right] \dot{ \theta_i } + K_D \Omega^2 \\+ K_D \dot{ \theta_i }^2 - \sum_{j \in \mathcal{N}_i} P_{ij}^{MAX} sin \left[ \theta_j - \theta_i \right] \end{array}
\end{equation}

These non linear equations are obviously not very practical. We now use some approximations based on the fact that we consider only small deviations from the main frequency : $ \dot{ \delta}_i \sim \Omega $ which means that $ \omega_i = \dot{\theta_i} << \Omega $, such that the squared term $ K_D \dot{\theta}_i^2 $ can be neglected.
Moreover, we assume that the rate at which energy is stored in the kinetic term is much less of the rate at which energy is dissipated by friction, a common condition for mechanical systems : $ \ddot{ \theta }_i  << \frac{2 K_D}{I} $

The equation can thus be simplified to :

@@

\begin{equation}
\ddot{\theta} = \Psi - \alpha \dot{\theta} - KL\theta
\end{equation}

@@

This is a continuous time second order linear system of $ N $ equations. We first transform this into a continuous time first order linear system of $ 2 N $ equations by introducing the following vector : $ X = \left( \begin{array}{c} \theta \\ \dot{\theta} \end{array} \right)$ :

@@

\begin{equation}
 Y(t+\Delta t) = A Y(t)
\end{equation}

@@

This is a linear discrete time system of $ 2 N + 1 $ equations. Note that the transition matrix A encodes all the system parameters, topology, and power distribution.

@@

This basically means that submodular functions exhibit a diminishing return property which makes them particularly interesting for optimization.

Finding the set $ S_k $ of size k that maximizes a set function F is a difficult problem because the number of sets grows exponentially with the number of nodes. Therefore complete enumeration and evaluation is only a feasible solution on very small examples. Nevertheless, if the set function is submodular, a simple greedy heuristic returns a solution $ S_k^{\star} $ such that, in the worst case, $ \frac{F(S_k^{\star})}{F(S_k^{OPT})} \sim 63\%$ (where $ S_k^{OPT}$ is the optimum set of size k) \cite{Krause2014}.

Since the evaluation of F can be costly, a well-known lazy-greedy variation has been proposed by \cite{Minoux}. This smart implementation uses the submodular structure of the marginal gains in order to reduce the number of calls to F. This can result in speedups of several order of magnitudes.

@@

We assume the control of a subset of the nodes that we call driver nodes. In our particular power grid situation, a driver node can be seen as a prosumer where some additional equipment (battery, backup generator,...) will be used as to inject or absorb energy from the system. This control assumption modifies the system dynamic :

@@

Where the control matrix B is an indicator of the driver nodes, and $ u(t) $ represents the control vector input at time t. The system is said to be controllable in T steps if it can be steered from any initial state $ Y_0 $ to any final state $ Y_f $ through a sequence of control inputs $ u(t) $. If the system is controllable, there might be more than one sequence of control inputs that could achieve such a thing. Among all these possibilities, we are interested in the one that requires the minimum amount of energy \cite{Yan2012}. Let $ \mathcal{E} = \int_{t_0}^{t_0+T} \Vert u(t) \Vert^2 dt $ be the energy used for the control of the system.

It can be shown \cite{Liu2015} that the control input sequence that minimizes the control energy can be written as :

@@

Not surprisingly, the control energy depends on the initial and final states as well as on the inverse of the Gramian. In our prosumer case, we want to find the best battery placement, whose quality should not depend on the initial and final states. Indeed, these states as well as  the power  distribution $ \Psi $ are susceptible to change, and we would like to perform well, on average, whatever the states and distributions. Therefore, we concentrate on  the Gramian as a way to quantify the average controllability. It can indeed be verified that the Gramian does not depend on $\Psi$.

A key point to design a low consumption control is thus to understand the relationships between the control node set and the Gramian matrix. In [], the authors introduce several Gramian-based metrics that have concrete meanings in terms of the system controllability. For example, the trace of the inverse Gramian ($ Tr[ W^{-1}(T)] $) quantifies the average control energy necessary to move the system around the state space. These metrics measures some notions of the control allowed by a given driver node set. But they can be understood the other way around : what drivers should one select as to optimize a given control-based metric ?

>>>>

Modernizing the power grid and increasing the share of renewables in the production are substantial objectives of the energetic transition. Because of progresses in communication, data management, and storage, the upcoming emergence of a power grid "2.0", often called smart grid, appears as a cross-disciplinary challenge of the 21st century \cite{Ramchurn}.

The today centralized top-down architecture is set to evolve to more distributed and bi-directional systems. Furthermore, the emergence of renewable and stochastic distributed energy resources (DER) in the distribution networks will require more flexibility. All these evolutions might change deeply the ways end users perceive and consume electricity. The consumer of the future, while seeking maximum utility/profit tradeoffs, is expected to be more involved in the system operation. Using dynamic signals such as real time electricity prices for instance, load curves could be shaped to some extend as to agree with the production conditions of the grid. Within these complex systems with numerous agents, it is assumed that multiple aggregation levels will be required in order to organize and optimize communication. Agents responsible for generation and load portfolio managment, i.e prosumers \cite{Rathnayaka2012}, will provide services (generation, load shedding, frequency regulation) against remuneration. Insure grid stability within this uncertain context appears as a complex task.

Several studies highlighted the need for increasing the storage capacity of the system. Some of them even consider electric vehicles as moving capacities that could be used for frequency regulation. In the case of fixed storage devices, the locations where they should be installed is an important question since it impacts the performances of the system. In a very static and top-down architecture, where loads and generators are fixed, this question is far from trivial. But in a smart grid scenario with bi-directional flows and agents that produce and consume depending on weather conditions, this becomes even more challenging. In this paper, we explore the use of storage in a network of prosumers whose power outputs are susceptible to change due to various external causes.

@@

More precisely, we model the power grid as a coupled oscillator network, where each node is represented as an oscillator with a phase angle and a natural frequency that are impacted by the ones of their neighbors. Since the work of Kuramoto [], networks of coupled oscillators have been used to model various phenomenons such as fireflies, ex2, or power grids. Indeed, it is well-known that synchronization of all rotating elements to a common frequency is necessary within a power grid for stable operation. Inspired by \cite{Filatrella2008}, we use a second order Kuramoto model in order to obtain a simplified dynamics for the power grid (see Section \ref{sec:The_Model}).

We will see that under some conditions, the system synchronizes to a common frequency $\Omega$. However, this synchronized state is suceptible to break if perturbations occur. If, for example, the production of some generator decreases while the load remains the same, the imbalance between production and consumption might steer the frequency away from $\Omega$ and drive the system to unstable conditions. These kinds of perturbations are very likely in a scenario with high penetration of DER. Multiple control techniques, with different time scales, exits in order to cope with such events. Among others, it has been argued that storage and vehicle to grid (V2G) technology could be used for frequency regulation. If a sudden drop or raise of the frequency due to an imbalance is detected, some devices are charged or discharged as to restore synchrony. This approach poses a few questions such as the location of the storage devices, or the quantities of power that should be injected or absorbed. Since charging and discharging a battery is costly, we are particularly interested in the energy necessary to restaure synchrony. Moreover, in a prosumer scenario where generators and loads tend not to be fixed, we would like to find a subset of locations for installing storage such that the energy is low on average.

We will see that these storage actions can be interpreted as an additional term in the oscillators dynamics for a subset of nodes. We propose here to use complex network control theory \cite{Liu2015} in order to model the impact of the storage equipments on the network dynamics (see Section \ref{sec:control_and_submodularity}). Placing storage at some node can thus be seen as selecting this node in the driver set, such that finding the locations for the storage devices could be related to finding the driver set of the network. Nevertheless, selecting the drivers that, on average, will control the network with low energy is a hard problem. Recent studies showed though the existence of relationships between controllability and submodular set functions \cite{Summers2014}. These functions, that we introduce in more depth in section X, have the interesting property of having diminishing returns as we increase the size of the set considered. Although finding the optimal set is NP-hard, \cite{Minoux} proposed a simple greedy algorithm with a worst case guaranty.

@@

The condition for the power transmission between i and j is that the two devices do not operate in phase : The phase difference between i and j is : $ \delta_j(t) - \delta_i(t) = \Omega t + \theta_j(t) - \Omega t - \theta_i(t) = \theta_j(t) - \theta_i(t) $. The transmitted power along the line can be written as :

@@

\begin{equation}
\label{eq:without_simplification}
\begin{array}{lll}
P_{S, i}  =  I \Omega \ddot{ \theta_i }  + \left[ I \ddot{ \theta_i } + 2 K_D \Omega \right] \dot{ \theta_i } + K_D \Omega^2 \\+ K_D \dot{ \theta_i }^2 - \sum_{j \in \mathcal{N}_i} P_{ij}^{MAX} sin \left[ \theta_j - \theta_i \right] \end{array}
\end{equation}

We now use simplifications based on the fact that we consider small deviations from the main frequency : $ \dot{ \delta}_i \sim \Omega $ which means that $ \omega_i = \dot{\theta_i} << \Omega $, such that the squared term $ K_D \dot{\theta}_i^2 $ can be neglected.
Moreover, we assume that the rate at which energy is stored in the kinetic term is much less of the rate at which energy is dissipated by friction : $ \ddot{ \theta }_i  << \frac{2 K_D}{I} $ (see \cite{Filatrella2008} for more details). Equation \ref{eq:without_simplification} becomes :

@@

\begin{equation}
\label{eq:with_simplification}
\ddot{\theta} = \Psi - \alpha \dot{\theta} - KL\theta
\end{equation}

@@

Equation \ref{eq:with_simplification} is a continuous time second order linear system of $ N $ equations. We first transform this into a continuous time first order linear system of $ 2 N $ equations by introducing the following vector : $ X = \left( \begin{array}{c} \theta \\ \dot{\theta} \end{array} \right)$ :

@@

\begin{equation}
\label{eq:final_equation}
 Y(t+\Delta t) = A Y(t)
\end{equation}

@@

Equation \ref{eq:final_equation} is a linear discrete time system of $ 2 N + 1 $ equations. Note that the transition matrix A encodes all the system parameters, topology, and power distribution.

@@

\begin{figure}
	\label{fig:submodular}
	\includegraphics[scale=.5]{submodular}
	\caption{Submodularity example with sensor placements.}
\end{figure}

This basically means that submodular functions exhibit a diminishing return property which makes them particularly interesting for optimization. A very intuitive example is the optimum placement of sensors in an area (see figure \ref{fig:submodular}). Sensors can be placed on a grid of locations and function F computes the surface of the space that is being sensed. Note first that adding a new sensor i to the current set S cannot decrease the value of F : $F(S+\{i\})>=F(S)$. Furthermore, if we add sensor i to a small set $S_1$ we tend to get larger improvements than if we add i to a larger set $S_2 \supset S_1$.

Finding the set $ S_k $ of size k that maximizes a set function F is a difficult problem because the number of sets grows exponentially with the number of nodes. Therefore complete enumeration and evaluation is only a feasible solution on very small examples. Nevertheless, if the set function is submodular, a simple greedy heuristic returns a solution $ S_k^{\star} $ such that, in the worst case, $ \frac{F(S_k^{\star})}{F(S_k^{OPT})} \sim 63\%$ (where $ S_k^{OPT}$ is the optimum set of size k) \cite{Krause2014}. This heuristic starts with a set S (possibly empty) and add the element i that exhibits the highest marginal gain : $ F(S \cup \{i\}) \geq F(S \cup \{j\})\ \forall j $.

For a ground set of N elements, this heuristic computes F $ \frac{k(2 N - k + 1)}{2} $ times. Since the evaluation of F can be costly, a well-known lazy-greedy variation has been proposed by \cite{Minoux}. This smart implementation uses the submodular structure of the marginal gains in order to reduce the number of calls to F. This requires to maintain a sorted table of marginal gains for all elements. When looking for new element i to add to set S, the top one is selected and the new marginal gain $F(S\cup \{§i\}) - F(S)$ is computed. If this gain is larger than the  gain of the second element in the table, then i is added to S. Otherwise i is inserted back in the table with its updated gain and the same treatment is applied to the element that is now on the top of the table. Because of the submodularity of F, this method performs as well as the original one, but can result in speedups of several order of magnitudes.

@@

We assume the control of a subset of the nodes that we call driver nodes. In our particular power grid situation, a driver node can be seen as a prosumer where some additional storage equipment is used to inject or absorb energy from the system. The dynamics of equation \ref{eq:final_equation} becomes :

@@

Where the control matrix B is an indicator of the driver nodes, and $ u(t) $ represents the control vector input at time t. The system is said to be controllable in T steps if it can be steered from any initial state $ Y_0 $ to any final state $ Y_f $ through a sequence of control inputs $ u(t) $. If the system is controllable, there might be more than one sequence of control inputs that could achieve such a thing. Among all these possibilities, we are interested in the one that requires the minimum amount of energy \cite{Yan2012}. Let $ \mathcal{E} = \int_{t_0}^{t_0+T} \Vert u(t) \Vert^2 dt $ be the energy used for the control of the system. It can be shown \cite{Liu2015} that the control input sequence that minimizes the control energy can be written as :

@@

Not surprisingly, the control energy depends on the initial and final states as well as on the inverse of the Gramian. In our prosumer case, we want to find the best battery placement, whose quality should not depend on the initial and final states. Indeed, these states as well as  the power  distribution $ \Psi $ are susceptible to change, and we would like to perform well, on average, whatever the states and distributions. Therefore, we concentrate on  the Gramian as a way to quantify the average controllability.

Since the control energy is related to $ W^{-1} $, small energies are obtained from "large" W. In [], the authors introduce several Gramian-based metrics that have concrete meanings in terms of the system controllability. For example, the trace of the inverse Gramian ($ Tr[ W^{-1}(T)] $) quantifies the average control energy necessary to move the system around the state space. These metrics measures some notions of the control allowed by a given driver node set. But they can be understood the other way around : what drivers should one select as to optimize a given control-based metric ?
